---
title: "Introduction to phylorug"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to phylorug}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```


## What problem does phylorug solve?

Phylogenetic trees represent best-supported, statistically convincing
hypotheses of evolutionary history, not ground truths (Steenwyk et al.
2023). Modern phylogenomic studies routinely produce multiple trees from
the same set of taxa. Different data types (UCEs, transcriptomes, whole
genomes), different analytical strategies (concatenation, coalescent
methods, site-heterogeneous models), and different inference software
(IQ-TREE, ASTRAL, MrBayes) each yield a tree with slightly different
topology and its own support values in its own format. At the same time,
no single tree inference method is proven to be the all-rounder that is
immune to all sources of error.

Incomplete lineage sorting, long-branch attraction, model
misspecification, and compositional heterogeneity each affect different
methods differently (Fleming et al. 2023). Accouringly, exploring
multiple strategeies , on different data types and different infrence
pipelines is the way researchers detect nodes that may be artifacts of a
particular analytical choice versus nodes that hold up regardless of
method.

The question at every node is:

Does this clade appear in all analyses, and how strongly does each one
support it?

Answering that by flipping between tree files is tedious, error-prone,
and does not scale. **phylorug** solves it by drawing a compact coloured
grid a **rug** at every internal node of a reference tree(backbone).
Each cell represents one tree, its colour shows whether that tree
recovered the clade and how strongly it supported it.

## Nodal support versus nodal stability

Giribet (2003) drew a distinction that motivates the two modes of
phylorug. **Nodal support** measures how confident a single analysis is
in a clade such as bootstrap, posterior probability, or LPP. **Nodal
stability** measures whether that clade is consistently recovered across
different analytical strategies, data types, and inference methods
,extending Wheeler's (1995) parameter-space sensitivity analysis
framework. These can decouple: a clade may receive 100% bootstrap under
one model yet collapse under all others, or carry only 30–50% support
everywhere yet appear in every tree.

phylorug shows both. In `presence` mode, each node gets a black dot if
every analysis recovers the clade (stable) or a rug showing which
analyses recover it and which do not.

In `support` mode, contested nodes receive a rug where each cell gives a
quantitative visual of the support value in that analysis , not just
presence or absence, but how strongly the clade is endorsed. Stable
nodes receive a black dot by default. Setting `rug_on_identical = TRUE`
extends the rug to these universally recovered nodes as well, revealing
cases where a clade is stable across all analyses but some analyses
support it weakly.

## Brief history

The rug plot concept originated with Wheeler (1995), who plotted clade
recovery across gap and transversion–transition cost ratios in parsimony
analysis under direct optimization (POY). Giribet (2003) named these
plots "Navajo rugs" and showed that nodal support and nodal stability
can tell different stories.

Sanders (2010) automated clade detection with Cladescan (Perl), and
Machado (2015) extended the approach with YBYRÁ (Python), which added
topological distance calculation and rogue taxon detection alongside
sensitivity analysis. Both tools produce individual SVG plots per node
that must be manually placed onto the tree in a vector editor. Cladescan
is no longer available online, and YBYRÁ has been archived by its
developer and is no longer maintained. More importantly, both were
designed around sensitivity analysis within a single analytical
framework — varying parameters and cost schemes — rather than comparing
trees from fundamentally different inference pipelines that report
support in incompatible formats (e.g., UFBoot2, SH-aLRT, posterior
probability, ASTRAL LPP).

phylorug extends this lineage into R. It reads trees from any pipeline,
bins support values against metric-specific thresholds rather than
rescaling them numerically, and draws rugs directly on the reference
tree — no manual placement, no external software.

## Quick start

The fastest way to see phylorug in action is with the built-in
`sample_trees` dataset — a 20-taxon subset of the Tarasov Lab beetle
phylogenomic data (Lopes et al., 2024), already rooted, pruned, and with
tip labels translated to species names.
```{r setup}
library(phylorug)
```

### Load the data

`sample_trees` is a named list of five `phylo` objects, all sharing the
same 20 taxa. Pick one as the backbone and use the rest as comparisons:

```{r load-sample}
names(sample_trees)

backbone <- sample_trees[["70p_uce"]]
others   <- sample_trees[names(sample_trees) != "70p_uce"]
```

### Check taxon consistency

Before building a rug, verify that every comparison tree carries the
same taxa as the backbone. `check_taxa()` reports the outcome and flags
any mismatches:

```{r check-taxa}
check_taxa(backbone, others)
```

All trees share the same 20 taxa, so we can proceed.

### Build the node presence matrix

`node_presence_matrix()` compares the backbone topology against each
comparison tree. For every internal node of the backbone, it records
whether the same clade (the same set of descendant tips) appears in each
comparison tree, and extracts any support values stored as node labels:

```{r build-npm}
npm <- node_presence_matrix(backbone, others, support_col = c(1, 2))
```

The result is a named list. `npm$presence` is a matrix of 1s (clade
recovered) and 0s (clade absent). `npm$support_1` and `npm$support_2`
hold the raw support values (first and second metric from compound node
labels like `"95/80"`), with `NA` where the clade was not recovered.

```{r inspect-npm}
npm$presence[1:5, ]
```

### Presence mode

The default plot shows presence/absence. Black cells mean the clade was
recovered; white cells mean it was not:

```{r quick-presence, fig.width = 6.2, fig.height = 5.8}
plot_phylorug(backbone, npm)
```

Nodes where all analyses agree get a solid black dot instead of a grid
(controlled by `dot_identical = TRUE`, the default). Grids appear only
where analyses disagree — these are the interesting nodes.

### Support mode

Switch to support mode to see how strongly each analysis supports each
clade. Cells are shaded by binned support strength (darker = stronger)
Support values are binned against thresholds specific to their own
metric. A tree with LPP 0.95 and a tree with LPP 0.65 are compared to
LPP thresholds, yielding different shades on the rug. A third tree
carrying UFBoot2 values is binned against UFBoot2 thresholds
independently — phylorug never cross-compares values from different
support metrics, because LPP 0.95 and UFBoot2 95 do not measure the same
thing despite looking numerically identical.

```{r quick-support,, fig.width = 6.2, fig.height = 5.8}
plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = c(
    "70p_partition_entropy"        = "ufboot",
    "70p_ghost"                    = "ufboot",
    "70p_ASTRAL_uce"               = "lpp",
    "70p_ASTRAL_partition_entropy" = "lpp"
  )
)
```

The `support_type` argument tells phylorug which support metric each
tree uses, so the correct thresholds are applied. IQ-TREE UFBoot2 values
are on a 0–100 scale; ASTRAL local posterior probabilities are on a 0–1
scale. phylorug normalises these automatically.

### Reading the plot: what you see at each node

Look at the output. Several distinct patterns appear at the nodes:

**Black dot** — the most common pattern on well-supported trees. All
comparison trees recover this clade unanimously. By default
(`dot_identical = TRUE`), phylorug draws a compact dot instead of a full
grid to reduce visual clutter.

**Mixed grid (black + white cells)** — these are the interesting nodes.
Some analyses recover the clade, others do not. The position legend at
the bottom tells you which cell corresponds to which analysis, so you
can see exactly which method or dataset disagrees.

**All-white grid** — no comparison tree recovers this backbone clade.
The clade is unique to the backbone topology. Set
`hide_unsupported = TRUE` to leave these nodes bare (no rug drawn at
all), which makes the remaining disagreements easier to read.

**Dot next to a rug** — adjacent nodes can differ in consensus. A deep
node might show a dot (all agree) while the node just above it shows a
mixed grid. This pattern often marks the boundary between well-resolved
and contentious regions of the tree.

Try `hide_unsupported = TRUE` to clean up the display:

```{r quick-hide, fig.width = 6.2, fig.height = 5.8}
plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = c(
    "70p_partition_entropy"        = "ufboot",
    "70p_ghost"                    = "ufboot",
    "70p_ASTRAL_uce"               = "lpp",
    "70p_ASTRAL_partition_entropy" = "lpp"
  ), 
  hide_unsupported = TRUE,
  cex = 1)
```

### Colour encoding in support mode

In **presence mode**, the encoding is binary: black (recovered) or white
(absent).

In **support mode**, cells use a greyscale gradient plus two special
colours:

- **Black / dark grey**: very high support (e.g., UFBoot \>= 95, LPP \>=
  0.95)
- **Medium grey**: high support
- **Light grey**: moderate support
- **Yellow**: low support (below moderate threshold, but clade is
  present)
- **White**: clade not recovered (absent)
- **Red**: not computed (the analysis did not evaluate this clade)

By default, unanimous nodes get a dot and no rug. To see the per-tree
support strength even at unanimous nodes, use `rug_on_identical = TRUE`:

```{r quick-rug-identical, fig.width = 6.2, fig.height = 5.8}
plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = c(
    "70p_partition_entropy"        = "ufboot",
    "70p_ghost"                    = "ufboot",
    "70p_ASTRAL_uce"               = "lpp",
    "70p_ASTRAL_partition_entropy" = "lpp"
  ),
  rug_on_identical = TRUE,
  cex = 1
)
```

## Working with real data: Culicomorpha

The built-in Culicomorpha dataset contains 10 phylogenomic analyses of
46 taxa from eight Diptera families (Fu et al., 2025), covering IQ-TREE
partitioned, PMSF, LG+C20+F+R, and ASTRAL strategies.

### Read the trees

`read_trees()` scans a directory for tree files and returns a named
list:

```{r read-culico}
tree_dir <- system.file("extdata", "culicomorpha", package = "phylorug")
trees    <- read_trees(tree_dir)
names(trees)
```

###Root and remove the outgroups

The Culicomorpha trees include three outgroup taxa from outside the
infraorder: Phlebotomus chinensis and Clogmia albipunctata (Psychodidae)
and Coboldia fuscipes (Scatopsidae). Root on all three, then drop them.
The order matters — you must root while the outgroup tips are still
present, because ape::root() needs to find them in the tree:

```{r root-culico}
trees <- lapply(trees, function(tr) {
  tr <- ape::root(tr, outgroup = c("Coboldia_fuscipes",
                                    "Phlebotomus_chinensis",
                                    "Clogmia_albipunctata"),
                  resolve.root = TRUE)
  ape::drop.tip(tr, c("Coboldia_fuscipes",
                       "Phlebotomus_chinensis",
                       "Clogmia_albipunctata"))
})
```

### Choose a backbone and check taxa

Pick one analysis as the backbone. The remaining trees are comparisons:

```{r backbone-culico}
backbone <- trees[["Matrix1-kpi_PMSF(H1.guide)"]]
others   <- trees[names(trees) != "Matrix1-kpi_PMSF(H1.guide)"]

check_taxa(backbone, others)
```

### Build the matrix and plot

```{r npm-culico}
npm <- node_presence_matrix(backbone, others)
```

```{r plot-culico, fig.width = 8, fig.height = 10}
plot_phylorug(backbone, npm)
```

With 9 comparison trees, phylorug automatically arranges the rug grid (3
columns by 3 rows in this case). The position legend at the bottom maps
each cell position to an analysis name.

```{r}
# --- Map each comparison tree to its support metric ---
support_type <- c(
  "Matrix1-kpi_ASTRAL"            = "lpp",
  "Matrix1-smart_ASTRAL"          = "lpp",
  "Matrix1-kpi_partitioning"      = "sh_alrt",
  "Matrix1-kpi_PMSF(H2.guide)"    = "sh_alrt",
  "Matrix1_LG_C20_F_R"            = "sh_alrt",
  "Matrix1-smart_partitioning"    = "sh_alrt",   # added — now a comparison tree
  "Matrix1-smart_PMSF(H1.guide)"  = "sh_alrt",
  "Matrix1-smart_PMSF(H2.guide)"  = "sh_alrt",
  "Matrix2_partitioning"          = "sh_alrt"
  # "Matrix1-kpi_PMSF(H1.guide)" removed — it's the backbone now, not a comparison
)
```

```{r plot-culico-support,fig.width = 8, fig.height = 10}
plot_phylorug(backbone, npm,
              mode               = "support",
              support_type       = support_type,
              show_support       = TRUE,
              support_label_cex  = 0.25,
              support_label_col  = "red",
              cell_scale         = 0.35,
              rug_position       = "inside",
              rug_on_identical   = FALSE,
              hide_unsupported   = TRUE)
```

## Full pipeline: beetles with tip translation

The beetle dataset in `inst/extdata/beetles_50p/` contains 5 analyses of
approximately 289 ingroup taxa, stored with specimen codes as tip labels
(e.g., `"OntauST002"`). A lookup table (`biogeo.csv`) maps these codes
to species names.

This example demonstrates the complete phylorug workflow, including tip
label translation.

### Read the trees

```{r read-beetles}
tree_dir <- system.file("extdata", "beetles_50p", package = "phylorug")
trees    <- read_trees(tree_dir)
names(trees)
```

### Root and remove outgroups

The beetle trees have two outgroup taxa: `NicorbUCE` and `NicvesUCE`.
Root on them, then drop them:

```{r root-beetles}
trees <- lapply(trees, function(tr) {
  tr <- ape::root(tr, outgroup = c("NicorbUCE", "NicvesUCE"),
                  resolve.root = TRUE)
  ape::drop.tip(tr, c("NicorbUCE", "NicvesUCE"))
})
```

### Translate tip labels

**Critical ordering rule:** root and remove the outgroup *before*
translating tip labels. Translation replaces the original specimen
codes, so if you translate first, `"NicorbUCE"` no longer exists and
`ape::root()` cannot find it.

Load the lookup table and translate:

```{r translate-beetles}
biogeo_path <- system.file("extdata", "beetles_50p", "biogeo.csv",
                           package = "phylorug")
biogeo <- read.csv(biogeo_path)

head(biogeo)

trees <- translate_tips(trees, biogeo,
                        from_col = "specimen_code",
                        to_col   = "species_name")
```

### Check taxa and build the matrix

```{r npm-beetles}
backbone <- trees[["50p_uce"]]
others   <- trees[names(trees) != "50p_uce"]

check_taxa(backbone, others)

npm <- node_presence_matrix(backbone, others, support_col = c(1, 2))
```

### Presence mode

```{r plot-beetles-presence, fig.width = 9, fig.height = 14}
plot_phylorug(backbone, npm, hide_unsupported = TRUE)
```

### Support mode

The 50p beetle trees use compound node labels (`"SH-aLRT/UFBoot2"` for
IQ-TREE, local posterior probability for ASTRAL). Map each tree to its
support metric:

```{r plot-beetles-support, fig.width = 9, fig.height = 14}
support_type <- c(
  "50p_partition_entropy"        = "ufboot",
  "50p_ghost"                    = "ufboot",
  "50p_ASTRAL_uce"               = "lpp",
  "50p_ASTRAL_partition_entropy" = "lpp"
)

plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = support_type,
  hide_unsupported = TRUE
)
```

## Customisation

The options below are demonstrated on the small `sample_trees` dataset
(20 taxa), where every detail is easy to inspect. Reload it so the
variables are clean:

```{r reload-sample}
backbone <- sample_trees[["70p_uce"]]
others   <- sample_trees[names(sample_trees) != "70p_uce"]
npm      <- node_presence_matrix(backbone, others, support_col = c(1, 2))

support_type <- c(
  "70p_partition_entropy"        = "ufboot",
  "70p_ghost"                    = "ufboot",
  "70p_ASTRAL_uce"               = "lpp",
  "70p_ASTRAL_partition_entropy" = "lpp"
)
```

### Hiding unsupported nodes

Some backbone clades are not recovered by any comparison tree, producing
all-white grids. Set `hide_unsupported = TRUE` to leave these nodes bare
— the absence of both a dot and a rug signals that the clade is unique
to the backbone:

```{r hide-unsupported, fig.width = 7, fig.height = 6}
plot_phylorug(backbone, npm, hide_unsupported = TRUE)
```

### Showing rugs on unanimous nodes

By default, unanimous nodes get a dot and no rug. To see per-tree
support strength even where every analysis agrees, add
`rug_on_identical = TRUE`:

```{r rug-on-identical, fig.width = 7, fig.height = 6}
plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = support_type,
  rug_on_identical = TRUE
)
```

### Cleanest figure: combining options

For the cleanest output — dots on unanimous nodes, rugs only where
analyses disagree, nothing where no analysis recovers the clade:

```{r cleanest, fig.width = 7, fig.height = 6}
plot_phylorug(backbone, npm, hide_unsupported = TRUE, dot_identical = TRUE)
```

### Grid dimensions

By default, phylorug chooses a roughly square grid. Override with
`n_rows` and `n_cols`:

```{r grid-demo, eval = FALSE}
plot_phylorug(backbone, npm, n_rows = 1, n_cols = 4)
plot_phylorug(backbone, npm, n_rows = 2, n_cols = 2)
```

### File output

For publication-quality figures, export to a file. The internal scaling
engine calculates optimal canvas dimensions automatically:

```{r file-output, eval = FALSE}
plot_phylorug(backbone, npm, file = "my_rug_plot.pdf")
plot_phylorug(backbone, npm, file = "my_rug_plot.png")
```

Override dimensions for journal requirements:

```{r file-custom, eval = FALSE}
plot_phylorug(backbone, npm, file = "figure_1.pdf",
              width = 12, height = 16)
```

### Tree appearance

`plot_phylorug()` passes additional arguments through `...` to
`ape::plot.phylo()`. Common options:

```{r tree-appearance, eval = FALSE}
# Tip label size, font style, branch width
plot_phylorug(backbone, npm, cex = 0.6, font = 3, edge.width = 1.5)
```

The backbone support labels (red numbers beside nodes) are controlled by
`show_support`, `support_label_cex`, and `support_label_col`:

```{r support-labels, eval = FALSE}
plot_phylorug(backbone, npm,
              show_support = TRUE,
              support_label_cex = 0.4,
              support_label_col = "black")
```

### Including the backbone

By default, the backbone does not occupy a cell in the rug (it trivially
recovers every clade, since it defines the topology). Set
`include_backbone = TRUE` to add it as cell 1:

```{r include-bb, eval = FALSE}
plot_phylorug(backbone, npm, include_backbone = TRUE)
```

In support mode with `include_backbone = TRUE`, you must also include
`"backbone"` in the `support_type` vector.

### Rug position

By default, rugs tuck into the crook above-left of each node
(`rug_position = "inside"`). Set `rug_position = "outside"` to place
them to the right of the node, toward the tips:

```{r rug-pos, eval = FALSE}
plot_phylorug(backbone, npm, rug_position = "outside")
```

### Custom thresholds

Override the default support bins with your own cutoffs:

```{r custom-thresh, eval = FALSE}
my_thresholds <- list(
  ufboot = c(very_high = 98, high = 90, moderate = 70),
  lpp    = c(very_high = 0.99, high = 0.90, moderate = 0.70)
)

plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = support_type,
  thresholds = my_thresholds
)
```

## Tips and best practices

**Root and prune before translating.** `translate_tips()` replaces
specimen codes with species names. Once translated, you cannot match
outgroup names for rooting. Always: (1) root, (2) drop outgroup, (3)
translate.

**One file = one analysis.** `read_trees()` treats each file as one
analysis. A file holding multiple equally optimal trees (as POY, TNT, or
PAUP\* write them) is scored as a pool — the presence value becomes the
proportion of pool trees recovering the clade. Do not pass posterior
samples or bootstrap replicates; those are distributions, not analyses.

**Wrap bare `multiPhylo` in a list.** If you construct comparison trees
manually, each element of the list passed to `node_presence_matrix()`
must be a single `phylo` or a `multiPhylo` (a pool from one search). A
bare `multiPhylo` at the top level is interpreted as one analysis with
multiple tied-optimal trees, not as separate analyses.

**Use `prune_to_shared()` when taxa differ.** If a comparison tree is
missing some backbone taxa, `node_presence_matrix()` will refuse to
proceed (a clade containing a missing taxon cannot be evaluated). Run
`check_taxa()` first to see the discrepancies, then `prune_to_shared()`
to reduce all trees to their common taxa.

## References

- Wheeler, W. C. (1995). Sequence alignment, parameter sensitivity, and
  the phylogenetic analysis of molecular data. *Systematic Biology*,
  44(3), 321–331.
- Sanders, K. L. (2010). Cladescan: exhaustive phylogenetic searches of
  consensus support in large data sets. *Cladistics*, 26(6), 598–613.
- Machado, D. J. (2015). YBYRÁ fossile: a command line system for total
  evidence dating and sensitivity analysis. *BMC Bioinformatics*, 16,
  40. 
- Lopes, F. et al. (2024). From museum drawer to tree: Historical DNA
  phylogenomics clarifies the systematics of rare dung beetles. *PLOS
  ONE*, 19(12), e0309596.
- Fu, Y. et al. (2025). Phylogenomic insights into the higher level
  relationships within Culicomorpha (Diptera). *Insect Systematics and
  Diversity*, 9(6), ixaf056.
