---
title: "Introduction to phylorug"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to phylorug}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

```{r setup}
library(phylorug)
```
## What problem does phylorug solve?

Modern phylogenomic studies routinely produce multiple trees from the same
set of taxa — different locus sets, alignment strategies, substitution models,
and inference methods each yield a tree that may agree or disagree with the
others. The question every systematist faces is: **which clades are robust
across analyses, and which are sensitive to analytical choices?**

Answering that question by flipping between tree files is tedious and
error-prone. phylorug solves it by drawing a small grid of coloured cells —
a **rug** — at every internal node of a single reference tree (the backbone).
Each cell in the grid represents one analysis, and its colour tells you whether
that analysis recovered the same clade, and how strongly it supported it.

The rug plot concept originates from Wheeler (1995), who first displayed
clade stability across analyses on a reference topology. Subsequent
implementations include Cladescan (Sanders, 2010) and YBYRÁ (Machado,
2015). phylorug brings this idea into R as a native package, with automatic
handling of heterogeneous support formats (UFBoot2, SH-aLRT, ASTRAL
local posterior probability, Bayesian posterior probability) across
independently run pipelines.

## Quick start

The fastest way to see phylorug in action is with the built-in
`sample_trees` dataset — a 20-taxon subset of the Tarasov Lab beetle
phylogenomic data (Lopes et al., 2024), already rooted, pruned, and
with tip labels translated to species names.

### Load the data

`sample_trees` is a named list of five `phylo` objects, all sharing the
same 20 taxa. Pick one as the backbone and use the rest as comparisons:

```{r load-sample}
names(sample_trees)

backbone <- sample_trees[["70p_uce"]]
others   <- sample_trees[names(sample_trees) != "70p_uce"]
```

### Check taxon consistency

Before building a rug, verify that every comparison tree carries the same
taxa as the backbone. `check_taxa()` reports the outcome and flags any
mismatches:

```{r check-taxa}
check_taxa(backbone, others)
```

All trees share the same 20 taxa, so we can proceed.

### Build the node presence matrix

`node_presence_matrix()` compares the backbone topology against each
comparison tree. For every internal node of the backbone, it records
whether the same clade (the same set of descendant tips) appears in each
comparison tree, and extracts any support values stored as node labels:

```{r build-npm}
npm <- node_presence_matrix(backbone, others, support_col = c(1, 2))
```

The result is a named list. `npm$presence` is a matrix of 1s (clade
recovered) and 0s (clade absent). `npm$support_1` and `npm$support_2`
hold the raw support values (first and second metric from compound node
labels like `"95/80"`), with `NA` where the clade was not recovered.

```{r inspect-npm}
npm$presence[1:5, ]
```

### Presence mode

The default plot shows presence/absence. Black cells mean the clade was
recovered; white cells mean it was not:

```{r quick-presence, fig.width = 15, fig.height = 9}
plot_phylorug(backbone, npm)
```

Nodes where all analyses agree get a solid black dot instead of a grid
(controlled by `dot_identical = TRUE`, the default). Grids appear only
where analyses disagree — these are the interesting nodes.

### Support mode

Switch to support mode to see how strongly each analysis supports each
clade. Cells are shaded by binned support strength (darker = stronger):

```{r quick-support, fig.width = 15, fig.height = 9}
plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = c(
    "70p_partition_entropy"        = "ufboot",
    "70p_ghost"                    = "ufboot",
    "70p_ASTRAL_uce"               = "lpp",
    "70p_ASTRAL_partition_entropy" = "lpp"
  )
)
```

The `support_type` argument tells phylorug which support metric each tree
uses, so the correct thresholds are applied. IQ-TREE UFBoot2 values are on
a 0–100 scale; ASTRAL local posterior probabilities are on a 0–1 scale.
phylorug normalises these automatically.

### Reading the plot: what you see at each node

Look at the output. Several distinct patterns appear at the nodes:

**Black dot** — the most common pattern on well-supported trees. All
comparison trees recover this clade unanimously. By default
(`dot_identical = TRUE`), phylorug draws a compact dot instead of a full
grid to reduce visual clutter.

**Mixed grid (black + white cells)** — these are the interesting nodes.
Some analyses recover the clade, others do not. The position legend at the
bottom tells you which cell corresponds to which analysis, so you can see
exactly which method or dataset disagrees.

**All-white grid** — no comparison tree recovers this backbone clade.
The clade is unique to the backbone topology. Set `hide_unsupported = TRUE`
to leave these nodes bare (no rug drawn at all), which makes the remaining
disagreements easier to read.

**Dot next to a rug** — adjacent nodes can differ in consensus. A deep
node might show a dot (all agree) while the node just above it shows a
mixed grid. This pattern often marks the boundary between well-resolved
and contentious regions of the tree.

Try `hide_unsupported = TRUE` to clean up the display:

```{r quick-hide, fig.width = 7, fig.height = 6}
plot_phylorug(backbone, npm, hide_unsupported = TRUE)
```

### Colour encoding in support mode

In **presence mode**, the encoding is binary: black (recovered) or
white (absent).

In **support mode**, cells use a greyscale gradient plus two special colours:

- **Black / dark grey**: very high support (e.g., UFBoot >= 95, LPP >= 0.95)
- **Medium grey**: high support
- **Light grey**: moderate support
- **Yellow**: low support (below moderate threshold, but clade is present)
- **White**: clade not recovered (absent)
- **Red**: not computed (the analysis did not evaluate this clade)

By default, unanimous nodes get a dot and no rug. To see the per-tree
support strength even at unanimous nodes, use `rug_on_identical = TRUE`:

```{r quick-rug-identical, fig.width = 7, fig.height = 6}
plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = c(
    "70p_partition_entropy"        = "ufboot",
    "70p_ghost"                    = "ufboot",
    "70p_ASTRAL_uce"               = "lpp",
    "70p_ASTRAL_partition_entropy" = "lpp"
  ),
  rug_on_identical = TRUE
)
```

## Working with real data: Culicomorpha

The built-in Culicomorpha dataset contains 10 phylogenomic analyses of
46 taxa from eight Diptera families (Fu et al., 2025), covering
IQ-TREE partitioned, PMSF, LG+C20+F+R, and ASTRAL strategies.

### Read the trees

`read_trees()` scans a directory for tree files and returns a named list:

```{r read-culico}
tree_dir <- system.file("extdata", "culicomorpha", package = "phylorug")
trees    <- read_trees(tree_dir)
names(trees)
```

### Root and remove the outgroup

These trees need rooting on *Pseudosmittia* sp., the outgroup. Root
first, then drop the outgroup tip. The order matters — you must root
while the outgroup is still present:

```{r root-culico}
trees <- lapply(trees, function(tr) {
  tr <- ape::root(tr, outgroup = "Pseudosmittia_sp", resolve.root = TRUE)
  ape::drop.tip(tr, "Pseudosmittia_sp")
})
```

### Choose a backbone and check taxa

Pick one analysis as the backbone. The remaining trees are comparisons:

```{r backbone-culico}
backbone <- trees[["Matrix1-kpi_partitioning"]]
others   <- trees[names(trees) != "Matrix1-kpi_partitioning"]

check_taxa(backbone, others)
```

### Build the matrix and plot

```{r npm-culico}
npm <- node_presence_matrix(backbone, others)
```

```{r plot-culico, fig.width = 8, fig.height = 10}
plot_phylorug(backbone, npm)
```

With 9 comparison trees, phylorug automatically arranges the rug grid
(3 columns by 3 rows in this case). The position legend at the bottom
maps each cell position to an analysis name.

## Full pipeline: beetles with tip translation

The beetle dataset in `inst/extdata/beetles_50p/` contains 5 analyses
of approximately 289 ingroup taxa, stored with specimen codes as tip
labels (e.g., `"OntauST002"`). A lookup table (`biogeo.csv`) maps these
codes to species names.

This example demonstrates the complete phylorug workflow, including
tip label translation.

### Read the trees

```{r read-beetles}
tree_dir <- system.file("extdata", "beetles_50p", package = "phylorug")
trees    <- read_trees(tree_dir)
names(trees)
```

### Root and remove outgroups

The beetle trees have two outgroup taxa: `NicorbUCE` and `NicvesUCE`.
Root on them, then drop them:

```{r root-beetles}
trees <- lapply(trees, function(tr) {
  tr <- ape::root(tr, outgroup = c("NicorbUCE", "NicvesUCE"),
                  resolve.root = TRUE)
  ape::drop.tip(tr, c("NicorbUCE", "NicvesUCE"))
})
```

### Translate tip labels

**Critical ordering rule:** root and remove the outgroup *before*
translating tip labels. Translation replaces the original specimen codes,
so if you translate first, `"NicorbUCE"` no longer exists and
`ape::root()` cannot find it.

Load the lookup table and translate:

```{r translate-beetles}
biogeo_path <- system.file("extdata", "beetles_50p", "biogeo.csv",
                           package = "phylorug")
biogeo <- read.csv(biogeo_path)

head(biogeo)

trees <- translate_tips(trees, biogeo,
                        from_col = "specimen_code",
                        to_col   = "species_name")
```

### Check taxa and build the matrix

```{r npm-beetles}
backbone <- trees[["50p_uce"]]
others   <- trees[names(trees) != "50p_uce"]

check_taxa(backbone, others)

npm <- node_presence_matrix(backbone, others, support_col = c(1, 2))
```

### Presence mode

```{r plot-beetles-presence, fig.width = 9, fig.height = 14}
plot_phylorug(backbone, npm, hide_unsupported = TRUE)
```

### Support mode

The 50p beetle trees use compound node labels (`"SH-aLRT/UFBoot2"` for
IQ-TREE, local posterior probability for ASTRAL). Map each tree to its
support metric:

```{r plot-beetles-support, fig.width = 9, fig.height = 14}
support_type <- c(
  "50p_partition_entropy"        = "ufboot",
  "50p_ghost"                    = "ufboot",
  "50p_ASTRAL_uce"               = "lpp",
  "50p_ASTRAL_partition_entropy" = "lpp"
)

plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = support_type,
  hide_unsupported = TRUE
)
```

## Customisation

The options below are demonstrated on the small `sample_trees` dataset
(20 taxa), where every detail is easy to inspect. Reload it so the
variables are clean:

```{r reload-sample}
backbone <- sample_trees[["70p_uce"]]
others   <- sample_trees[names(sample_trees) != "70p_uce"]
npm      <- node_presence_matrix(backbone, others, support_col = c(1, 2))

support_type <- c(
  "70p_partition_entropy"        = "ufboot",
  "70p_ghost"                    = "ufboot",
  "70p_ASTRAL_uce"               = "lpp",
  "70p_ASTRAL_partition_entropy" = "lpp"
)
```

### Hiding unsupported nodes

Some backbone clades are not recovered by any comparison tree, producing
all-white grids. Set `hide_unsupported = TRUE` to leave these nodes
bare — the absence of both a dot and a rug signals that the clade is
unique to the backbone:

```{r hide-unsupported, fig.width = 7, fig.height = 6}
plot_phylorug(backbone, npm, hide_unsupported = TRUE)
```

### Showing rugs on unanimous nodes

By default, unanimous nodes get a dot and no rug. To see per-tree
support strength even where every analysis agrees, add
`rug_on_identical = TRUE`:

```{r rug-on-identical, fig.width = 7, fig.height = 6}
plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = support_type,
  rug_on_identical = TRUE
)
```

### Cleanest figure: combining options

For the cleanest output — dots on unanimous nodes, rugs only where
analyses disagree, nothing where no analysis recovers the clade:

```{r cleanest, fig.width = 7, fig.height = 6}
plot_phylorug(backbone, npm, hide_unsupported = TRUE, dot_identical = TRUE)
```

### Grid dimensions

By default, phylorug chooses a roughly square grid. Override with
`n_rows` and `n_cols`:

```{r grid-demo, eval = FALSE}
plot_phylorug(backbone, npm, n_rows = 1, n_cols = 4)
plot_phylorug(backbone, npm, n_rows = 2, n_cols = 2)
```

### File output

For publication-quality figures, export to a file. The internal scaling
engine calculates optimal canvas dimensions automatically:

```{r file-output, eval = FALSE}
plot_phylorug(backbone, npm, file = "my_rug_plot.pdf")
plot_phylorug(backbone, npm, file = "my_rug_plot.png")
```

Override dimensions for journal requirements:

```{r file-custom, eval = FALSE}
plot_phylorug(backbone, npm, file = "figure_1.pdf",
              width = 12, height = 16)
```

### Tree appearance

`plot_phylorug()` passes additional arguments through `...` to
`ape::plot.phylo()`. Common options:

```{r tree-appearance, eval = FALSE}
# Tip label size, font style, branch width
plot_phylorug(backbone, npm, cex = 0.6, font = 3, edge.width = 1.5)
```

The backbone support labels (red numbers beside nodes) are controlled by
`show_support`, `support_label_cex`, and `support_label_col`:

```{r support-labels, eval = FALSE}
plot_phylorug(backbone, npm,
              show_support = TRUE,
              support_label_cex = 0.4,
              support_label_col = "black")
```

### Including the backbone

By default, the backbone does not occupy a cell in the rug (it trivially
recovers every clade, since it defines the topology). Set
`include_backbone = TRUE` to add it as cell 1:

```{r include-bb, eval = FALSE}
plot_phylorug(backbone, npm, include_backbone = TRUE)
```

In support mode with `include_backbone = TRUE`, you must also include
`"backbone"` in the `support_type` vector.

### Rug position

By default, rugs tuck into the crook above-left of each node
(`rug_position = "inside"`). Set `rug_position = "outside"` to place
them to the right of the node, toward the tips:

```{r rug-pos, eval = FALSE}
plot_phylorug(backbone, npm, rug_position = "outside")
```

### Custom thresholds

Override the default support bins with your own cutoffs:

```{r custom-thresh, eval = FALSE}
my_thresholds <- list(
  ufboot = c(very_high = 98, high = 90, moderate = 70),
  lpp    = c(very_high = 0.99, high = 0.90, moderate = 0.70)
)

plot_phylorug(
  backbone, npm,
  mode = "support",
  support_type = support_type,
  thresholds = my_thresholds
)
```

## Tips and best practices

**Root and prune before translating.** `translate_tips()` replaces
specimen codes with species names. Once translated, you cannot match
outgroup names for rooting. Always: (1) root, (2) drop outgroup,
(3) translate.

**One file = one analysis.** `read_trees()` treats each file as one
analysis. A file holding multiple equally optimal trees (as POY, TNT,
or PAUP\* write them) is scored as a pool — the presence value becomes
the proportion of pool trees recovering the clade. Do not pass posterior
samples or bootstrap replicates; those are distributions, not analyses.

**Wrap bare `multiPhylo` in a list.** If you construct comparison trees
manually, each element of the list passed to `node_presence_matrix()`
must be a single `phylo` or a `multiPhylo` (a pool from one search).
A bare `multiPhylo` at the top level is interpreted as one analysis with
multiple tied-optimal trees, not as separate analyses.

**Use `prune_to_shared()` when taxa differ.** If a comparison tree is
missing some backbone taxa, `node_presence_matrix()` will refuse to
proceed (a clade containing a missing taxon cannot be evaluated).
Run `check_taxa()` first to see the discrepancies, then
`prune_to_shared()` to reduce all trees to their common taxa.

## References

- Wheeler, W. C. (1995). Sequence alignment, parameter sensitivity,
  and the phylogenetic analysis of molecular data. *Systematic Biology*,
  44(3), 321–331.
- Sanders, K. L. (2010). Cladescan: exhaustive phylogenetic searches
  of consensus support in large data sets.
  *Cladistics*, 26(6), 598–613.
- Machado, D. J. (2015). YBYRÁ fossile: a command line system for
  total evidence dating and sensitivity analysis.
  *BMC Bioinformatics*, 16, 40.
- Lopes, F. et al. (2024). From museum drawer to tree: Historical DNA
  phylogenomics clarifies the systematics of rare dung beetles.
  *PLOS ONE*, 19(12), e0309596.
- Fu, Y. et al. (2025). Phylogenomic insights into the higher level
  relationships within Culicomorpha (Diptera).
  *Insect Systematics and Diversity*, 9(6), ixaf056.















