---
title: "Introduction to phylorug"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to phylorug}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

# What problem does phylorug solve?

Modern phylogenomic studies routinely produce multiple trees from
the same set of taxa. Different data types (UCEs, transcriptomes, whole
genomes), different analytical strategies (concatenation, coalescent
methods, site-heterogeneous models), and different inference software
(IQ-TREE, ASTRAL, MrBayes) each yield a tree with slightly different
topology and its own support values in its own format. At the same time,
no single tree inference method is proven to be the all-rounder.

Incomplete lineage sorting, long-branch attraction, model
misspecification, and compositional heterogeneity each affect different
methods differently (Fleming et al. 2023). Accordingly, exploring
multiple strategies , on different data types and different inference
pipelines is the way researchers detect nodes that may be artifacts of a
particular analytical choice versus nodes that hold up regardless of
method.

The question at every node is:

Does this clade appear in all analyses, and how strongly does each one
support it?

Answering that by flipping between tree files is tedious, error-prone,
and is very time-consuming with bigger trees. **phylorug** solves it by drawing a compact coloured grid a **rug plot** at every internal node of a reference tree(backbone).
Each cell represents one tree, its colour shows whether that tree
recovered the clade and how strongly it supported it.

# Nodal support versus nodal stability

Giribet (2003) drew a distinction that motivates the two modes of
phylorug. **Nodal support** measures how confident a single analysis is
in a clade such as bootstrap, posterior probability, or LPP. **Nodal
stability** measures whether that clade is consistently recovered across
different analytical strategies, data types, and inference methods,
extending Wheeler's (1995) parameter-space sensitivity analysis
framework. These can decouple: a clade may receive 100% bootstrap under
one model yet collapse under all others, or carry only 30–50% support
everywhere yet appear in every tree.

phylorug shows both. In `presence` mode, each node gets a black dot if
every analysis recovers the clade (stable), or a rug showing which
analyses recover it and which do not — presence or absence only.

`support` mode builds on that: it keeps the same presence/absence
information but adds how strongly each analysis supports the clade,
shading each cell by its support value instead of just marking it
filled or empty. Stable nodes still get a black dot by default. Setting
`rug_on_identical = TRUE` extends the rug to these nodes too — useful in
support mode, since it shows whether a clade recovered by every
analysis is actually supported equally strongly by all of them.

## Brief history

The rug plot concept originated with Wheeler (1995), who plotted clade
recovery across gap and transversion–transition cost ratios in parsimony
analysis. Giribet (2003) named these plots "Navajo rugs" and showed that
nodal support and nodal stability can tell different stories.

Sanders (2010) automated parameter-space sensitivity analysis with Cladescan 
(Perl), and Machado (2015) extended the approach with YBYRÁ (Python). Both tools
produce individual SVG plots per node that must be manually placed onto the tree
in a vector editor. Cladescan is no longer available online, and YBYRÁ has been 
archived by its developer and is no longer maintained. More importantly, both were designed around sensitivity analysis within a single analytical framework, varying parameters and cost schemes, rather than comparing trees from fundamentally different inference pipelines that report support in incompatible formats (e.g.,UFBoot2, SH-aLRT, posterior probability, ASTRAL LPP).

phylorug extends this lineage into R. It reads trees from any pipeline,
bins support values against thresholds built for their own
metric instead of rescaling everything onto one numeric scale, and
draws the rug directly on the reference tree, no manual placement, no
outside software.

# Quick start

The fastest way to see phylorug in action is with the built-in
`sample_trees` dataset, a 15-taxon subset of the beetle data (Montanaro, 
Lopes et al., 2026), already rooted, pruned, and with tip labels translated to
species names. Lets start with attaching the package :

```{r setup}
library(phylorug)
```

## Load the data

`sample_trees` is a named list of five `phylo` objects, all sharing the
same 15 taxa. Pick one as the backbone and use the rest as comparisons.
In the original study `70p_uce` is used as a backbone tree.

```{r load-sample}
names(sample_trees)

backbone <- sample_trees[["70p_uce"]]
others   <- sample_trees[names(sample_trees) != "70p_uce"]
```

## Check taxon consistency

`check_taxa()` compares the tip labels of the backbone against every
comparison tree and reports any mismatches , missing taxa, extra taxa,
or spelling differences. Running it before `node_presence_matrix()` is
good practice, especially the first time you work with a new dataset.
That said, `node_presence_matrix()` will stop with a clear error if a
comparison tree is missing a backbone taxon, so if you already know your
trees share the same taxa you can skip straight to building the matrix :

```{r check-taxa}
check_taxa(backbone, others)
```

All trees share the same 15 taxa, so we can proceed. IF a mismatch is
reported the helper function `prune_to_shared()` can be used to get
identical taxa set.

## Build the node presence matrix

Once the trees share the same taxa, the next question is: which clades
from the backbone appear in which comparison trees, and how strongly are
they supported? `node_presence_matrix()` answers this by walking every
internal node of the backbone and checking each comparison tree for the
same clade.

It returns a named list of matrices: a **presence matrix** (1 = clade
recovered, 0 = absent) and one or more **support matrices** containing
the support values extracted from the node labels of each tree. By
default, the first support value in each node label is used. If your
trees carry compound labels like `100/98` (e.g., SH-aLRT/UFBoot2 from
IQ-TREE), `support_col` lets you select which value to extract,
`support_col = 1` takes the first, `support_col = 2` takes the second.

Two other arguments belong here rather than at plot time. `support_type`
is an optional named character vector that declares which support metric
each comparison tree uses (e.g., `"ufboot"`, `"lpp"`, `"sh_alrt"`,
`"posterior"`). It is stored as an attribute of the result and read
automatically by `plot_phylorug()` when drawing in support mode, so you
only declare it once.

When a comparison tree file contains multiple equally optimal trees (a
pool), `pool_threshold` decides how many of them must recover a clade
for it to count as present. At `1.0` (the default), all of them must be
a strict consensus. At `0.5`, a majority is enough. At `0`, the cell
simply shows the fraction that recovered it, with no cutoff applied. The
`sample_trees` dataset contains single `phylo` objects, not pools, so
`pool_threshold` does not apply here. We declare support_type so that
`plot_phylorug()` can bin each tree's support values against the correct
metric-specific thresholds:

```{r Support_type_vector}
support_type <- c(
  "70p_ASTRAL_partition_entropy" = "lpp",
  "70p_ASTRAL_uce"               = "lpp",
  "70p_ghost"                    = "ufboot",
  "70p_partition_entropy"        = "ufboot"
)
```

```{r build-npm}
npm <- node_presence_matrix(backbone, others, support_col = c(1, 2), support_type = support_type)
```

The result is a named list. `npm$presence` is a matrix of 1s and 0s.
`npm$support_1` and `npm$support_2` hold the raw support values, with
`NA` where the clade was not recovered. These matrices are what
`plot_phylorug()` uses to draw the rug, presence determines whether a
cell is filled or empty, support determines its shade.

## Plot the rug

With the node presence matrix ready, `plot_phylorug()` draws the rug on
the backbone. In its simplest form, all arguments are defaults:
`mode = "presence"`, `dot_identical = TRUE` (unanimous nodes get a black
dot), `legend = TRUE`, and `rug_position = "inside"`. The internal
scaling engine sizes the canvas, fonts, and cell dimensions
automatically based on tree size.

```{r quick-presence,  fig.height=5.8, fig.width= 6.5, fig.dpi=300}
plot_phylorug(backbone, npm)
```

For publication-quality output, setting `file`, `width`, and `height` is
strongly recommended. All other arguments are worth exploring to
fine-tune your figure, but these three matter most.

```{r presence-file, eval = FALSE}
plot_phylorug(backbone, npm, file = "presence.pdf", width = 12, height = 8)
```

## Support mode

Now switch to support mode to see how strongly each tree supports each
clade. Cells are shaded by binned support strength, darker means
stronger. Support values are binned against thresholds specific to their
own metric: a tree with LPP 0.95 and a tree with LPP 0.65 are compared
to LPP thresholds, while a tree carrying UFBoot2 values is binned
against UFBoot2 thresholds independently. phylorug never cross-compares
values from different support metrics, because LPP 0.95 and UFBoot2 95
do not measure the same thing despite looking numerically similar.

Since we already declared `support_type` when building the node presence
matrix, `plot_phylorug()` reads it automatically from the stored
attribute and applies the correct metric-specific thresholds.

If `support_type` is not declared, `plot_phylorug()` falls back to a
universal threshold scale: values greater than 1 are divided by 100 to
bring them onto a 0–1 scale, and all trees are binned against the same
set of thresholds regardless of metric. This is a reasonable starting
point for quick exploration, but declaring `support_type` is always
preferred because it ensures each value is evaluated against the
thresholds published for its own method.

```{r support-default, fig.height=5.8, fig.width= 6.5, fig.dpi=300}
plot_phylorug(
  backbone, npm,
  mode         = "support")
```

# Reading the rug plot

The rug plot carries two legends. The **position legend** (top-left)
maps each numbered cell to a tree. For example, cell 1 is the first
tree, cell 2 the second, and so on. The **threshold legend** (top-right,
support mode only) shows what each shade means.

At each node you will see one of these patterns:

- **Black dot**: all analyses recover this clade unanimously
  (`dot_identical = TRUE`, the default). No rug is drawn to reduce clutter.
- **Mixed grid**: some cells filled with solid black, some white. These
  are the interesting nodes, the position legend tells you which analysis agrees  and which disagrees.
- **Bare node** (no dot, no rug): no comparison tree recovers this backbone 
  clade, it exists only in the backbone topology. This is the default behaviour    (`hide_unsupported = TRUE`). Set `hide_unsupported = FALSE` to draw all-white    grids at these nodes instead.

In support mode, filled cells are shaded by how strongly each analysis supports the clade. The colour scheme uses a greyscale gradient for recovered clades plus two special colours for absent and unparseable cases. The thresholds shown below are the package defaults. Different fields and journals apply different cutoffs, so we recommend defining your own via the `thresholds` argument (see the Customisation section below):

- **Black**: very high support (UFBoot2 \>= 95, SH-aLRT \>= 80, LPP \>=0.95).
- **Dark grey**: high support (UFBoot2 80-94, SH-aLRT 70-79, LPP 0.90-0.94).
- **Light grey**: moderate support (UFBoot2 50-79, SH-aLRT 50-69, LPP 0.50-0.89).
- **Yellow**: low support (below 50 for UFBoot2/SH-aLRT, below 0.50 for LPP). Present but weakly endorsed.
- **White**: clade not recovered by that analysis.
- **Red**: clade recovered but no support value could be read. This typically means the tree file had no node labels, or the label was not a parseable number.

# Fine-tuning the figure

The following example shows how optional arguments refine the plot.
`rug_on_identical = TRUE` extends the rug to unanimous nodes, revealing
per-tree support strength even at stable clades.
`hide_unsupported = FALSE` would draw all-white rugs at nodes unique to the backbone; the default (`TRUE`) leaves them bare.
`cell_scale` adjusts the size of each rug cell, `rug_position` controls
whether rugs sit toward the root (`"inside"`) or toward the tips
(`"outside"`). `dot_cex` scales the unanimous-node dot, and
`show_support = TRUE` with `support_label_cex` and `support_label_col`
overlays the backbone's own support labels on the tree. `cex` and `font`
are passed directly to the tree plotter. For publication output, add
`file = "output.pdf"` along with `width` and `height` to produce a
properly scaled figure.

```{r support-refined, fig.height=5.8, fig.width= 6.5, fig.dpi=300}
# For publication output, add:
# file = "my_support_plot.pdf"
plot_phylorug(
  backbone, npm,
  width             = 6.5,
  height            = 5.8,
  mode              = "support",
  rug_position      = "inside",
  cell_scale        = 0.38,
  rug_on_identical  = TRUE,
  cex               = 0.90,
  font              = 3,
  dot_cex           = 1.3,
  show_support      = TRUE,
  support_label_col = "red",
  support_label_cex = 0.4
)
```

# Working with real data: Culicomorpha

The built-in Culicomorpha dataset contains 10 phylogenomic analyses of
46 taxa from eight Diptera families (Fu et al., 2025), covering IQ-TREE
partitioned, PMSF, LG+C20+F+R, and ASTRAL strategies.

## Read the trees

`read_trees()` scans a directory for tree files and returns a named
list:

```{r read-culico}
tree_dir <- system.file("extdata", "culicomorpha", package = "phylorug")
trees    <- read_trees(tree_dir)
names(trees)
```

## Root and remove the outgroups

The *Culicomorpha* trees include three outgroup taxa from outside the
infraorder: *Phlebotomus chinensis* and *Clogmia albipunctata
(Psychodidae)* and *Coboldia fuscipes (Scatopsidae)*. Root on all three,
then drop them. The order matters — you must root while the outgroup
tips are still present, because `ape::root()` needs to find them in the
tree:

```{r root-culico}
trees <- lapply(trees, function(tr) {
  tr <- ape::root(tr, outgroup = c("Coboldia_fuscipes",
                                   "Phlebotomus_chinensis",
                                   "Clogmia_albipunctata"),
                  resolve.root = TRUE)
  ape::drop.tip(tr, c("Coboldia_fuscipes",
                      "Phlebotomus_chinensis",
                      "Clogmia_albipunctata"))
})
```

The Culicomorpha trees already carry species names as tip labels, so no
translation is needed here. For trees with specimen codes, see
`translate_tips()` in the beetles example below.

## Choose a backbone

Pick one analysis as the backbone. The remaining trees are comparisons:

```{r backbone-culico}
backbone <- trees[["Matrix1-kpi_PMSF(H1.guide)"]]
others   <- trees[names(trees) != "Matrix1-kpi_PMSF(H1.guide)"]

```

## Check taxon consistency

```{r check-taxon-consistency-culico}
check_taxa(backbone, others)
```

## Build the matrix and plot
```{r support-type-culico}
support_type <- c(
  "Matrix1-kpi_ASTRAL"            = "lpp",
  "Matrix1-smart_ASTRAL"          = "lpp",
  "Matrix1-kpi_partitioning"      = "sh_alrt",
  "Matrix1-kpi_PMSF(H2.guide)"    = "sh_alrt",
  "Matrix1_LG_C20_F_R"            = "sh_alrt",
  "Matrix1-smart_partitioning"    = "sh_alrt",
  "Matrix1-smart_PMSF(H1.guide)"  = "sh_alrt",
  "Matrix1-smart_PMSF(H2.guide)"  = "sh_alrt",
  "Matrix2_partitioning"          = "sh_alrt"
)
```

```{r npm-culico}
npm <- node_presence_matrix(backbone, others, support_type = support_type)
```

## plot the phylorug-presence

```{r plot-culico, fig.width = 8, fig.height = 10}
plot_phylorug(backbone, npm)
```

With 9 comparison trees, phylorug automatically arranges the rug grid (3
columns by 3 rows in this case). The position legend maps each cell
position to an analysis name.

## support-mode

Here we use show_support = TRUE to overlay the backbone's own node labels in red for cross-referencing and `cell_scale = 0.35` to shrink the rug cells slightly for a denser tree. All-white rugs are hidden by default, so only contested and stable nodes remain visible.

```{r plot-culico-support,fig.width = 8, fig.height = 10}
plot_phylorug(backbone, npm,
              mode               = "support",
              show_support       = TRUE,
              support_label_cex  = 0.25,
              support_label_col  = "red",
              cell_scale         = 0.35,
              rug_position       = "inside",
              rug_on_identical   = FALSE)
```

## What the rug reveals?

The **rugs** immediately separates stable from contested regions of the
*Culicomorpha* tree. Most family-level clades carry black dots, meaning
those clades are recovered by unanimously all 9 analyses. *Culicidae*
(mosquitoes), *Chironomidae* (non-biting midges), *Ceratopogonidae*
(biting midges), and the *Simuliidae* + *Thaumaleidae* clade all show
complete agreement across every inference method and dataset. Fu et al.
(2025) reported the same, full support for these clades regardless of
model or matrix.

The interesting nodes are the contested ones. The central question in
*Culicomorpha* phylogenetics is where *Ceratopogonidae* and
*Chironomidae* sit relative to the rest of the infraorder. Fu et al.
(2025) tested multiple hypotheses. Their preferred topology (H1) places
*Chironomidae* + *Ceratopogonidae* together as sister to all remaining
families. The alternative (H2) breaks this pairing and places
*Ceratopogonidae* elsewhere, nested closer to the other families. On the
rug, the node defining the *Chironomidae* + *Ceratopogonidae* clade
shows a mixed grid. The two ASTRAL cells (cells 1 and 4 in the position
legend) are white at this node, meaning the coalescent-based analyses
did not recover this grouping. The IQ-TREE concatenation cells are
mostly black, meaning the concatenation analyses did recover it. This is
a textbook example of gene-tree/species-tree conflict made visible on a
single figure: concatenation says these two families group together,
coalescent methods say they do not.

Within *Chironomidae*, the subfamily-level relationships are largely
stable, with black dots on most internal nodes. A few nodes near
*Potthastia* and *Paraheptagyia* show mixed grids with red cells. Red
means the analysis recovered the clade but carried no computable support
value. These nodes sit on short internal branches where rapid
diversification left little phylogenetic signal, making resolution
sensitive to model choice. Fu et al. (2025) noted similar instability
around the placement of *Telmatogetoninae* within *Diamesinae* depending
on the analytical model used.

This is the core value of `phylorug`: patterns that required opening 10
separate tree files side by side in the original study are condensed
onto a single figure. A reader can immediately see which nodes are
robust to analytical choice and which deserve further investigation.

# Full pipeline: beetles with tip translation

The beetle dataset ships with `phylorug` in `inst/extdata/`. It contains
two sets of trees, a random subset of Montanaro, Lopes et al. (2026): beetles_50p/ and beetles_70p/, each with 5 analyses of 50 ingroup dung beetle taxa (52 tips including 2 outgroups) for the 50p set, and 70 ingroup taxa (72 tips) for the 70p set built from UCE loci retained at different completeness thresholds (50% and 70%). The two datasets
produce slightly different topologies, making them a good test case for
exploring how data filtering affects clade recovery. For this vignette
we use the 50p set, but users are encouraged to try both. Unlike the
*Culicomorpha* trees, these trees store specimen codes as tip labels
(e.g., `"OntauST002"` rather than a species name), so this example adds
one extra step: translating tip labels using a lookup table before
building the rug.

## Read the trees

```{r read trees-beetles-50p}
tree_dir <- system.file("extdata", "beetles_50p", package = "phylorug")
trees    <- read_trees(tree_dir)
names(trees)
```

## Root and remove outgroups

The beetle trees include two outgroup taxa, *NicorbUCE* and *NicvesUCE*.
We ill root on them first, then drop them. As before, rooting must
happen while the outgroup tips are still in the tree.

```{r root and remove outgroup- bettle-50p}
trees <- lapply(trees, function(tr) {
  tr <- ape::root(tr, outgroup = c("NicorbUCE", "NicvesUCE"),
                  resolve.root = TRUE)
  ape::drop.tip(tr, c("NicorbUCE", "NicvesUCE"))
})
```

## Translate tip labels

The trees still carry specimen codes at this point. A CSV file
`biogeo.csv` included with the package maps each code to its species
name. Lets load it and check the first few rows:

```{r load the lookup table}
biogeo_path <- system.file("extdata", "beetles_50p", "biogeo.csv",
                           package = "phylorug")
biogeo <- read.csv(biogeo_path)
head(biogeo)
```

Now translate. The ordering rule from earlier applies here. Frist root
and drop outgroups before translating, because `translate_tips()`
replaces the original codes and `ape::root()` would no longer find
"*NicorbUCE*" after translation.

```{r translate tree tips-beetle-50p}
trees <- translate_tips(trees, biogeo,
                        from_col = "specimen_code",
                        to_col   = "species_name")
```

## Choose a backbone and check taxa

```{r Choose a backbone and check taxa}
backbone <- trees[["50p_uce"]]
others   <- trees[names(trees) != "50p_uce"]
check_taxa(backbone, others)
```

## Build the node presence matrix

The beetle trees carry compound node labels (`SH-aLRT/UFBoot2` for
`IQ-TREE trees`). Extract both support values with
`support_col = c(1, 2)`:
```{r support_type_define}
support_type <- c(
  "50p_partition_entropy"        = "ufboot",
  "50p_ghost"                    = "ufboot",
  "50p_ASTRAL_uce"               = "lpp",
  "50p_ASTRAL_partition_entropy" = "lpp"
)
```

```{r npm-bettle-50p}
npm <- node_presence_matrix(backbone, others, support_col = c(1, 2),
                            support_type = support_type)
```

## Presence mode

At \~52 taxa the tree is still comfortably readable. All-white rugs are
hidden by default, and writing to a file with explicit width and height
avoids the distortion that GUI windows introduce.

```{r plot phylorug , fig.height= 12, fig.width= 10 }
plot_phylorug(backbone, npm)
```

## Support mode
With ~52 taxa, the canvas still comfortably fits a single page. `show_support = TRUE` with `support_label_col = "red"` overlays the backbone's own support values for cross-referencing against the **rug** shading, and `cell_scale = 0.35` keeps the rug cells from crowding each other.

```{r plot phylorug beetles 50 p, fig.height= 12, fig.width= 10}
# For publication output, add:
# file = "beetles_support.pdf"
plot_phylorug(
  backbone,
  npm,
  mode               = "support",
  show_support       = TRUE,
  support_label_col  = "red",
  support_label_cex  = 0.34,
  cell_scale         = 0.35,
  rug_on_identical   = FALSE,
  cex                = 0.8
)
```

## What the rug reveals?
Even at 50 taxa, the rug lines up with real structure from the source
study. The *Helictopleurus* clade sits beside *Onthophagus*,
*Proagoderus*, *Cheironitis*, and *Megalonitis*  together these form
Onthophagini *sensu novo*, the tribe Montanaro, Lopes et al. (2026)
redefined to merge the former Onthophagini and Oniticellini, with
*Helictopleurus* placed in the newly established subtribe Helictopleurina.

Three nodes disagree across the four analyses. The clearest case is the
node uniting *Helictopleurus semivirens* and *H. undatus* (backbone
support 82.8): both ASTRAL trees fail to recover it, while both
concatenation trees (GHOST and the partitioned IQ-TREE run) do — a
clean coalescent-versus-concatenation split. The other two disagreements
are messier. At the node uniting *Nanos sp1* and *N.* aff. *Bicoloratus*
(backbone support 100.0), only the ASTRAL UCE tree recovers the clade;
the other three analyses, including the other ASTRAL run, do not. That
lines up with what the source study itself flags: the *Nanos* generic
group is left unresolved on purpose, its placement reserved for a
forthcoming study.

At the node uniting *Helictopleurus sicardi* and *H. viettei* (backbone
support 100.0), only one analysis (ASTRAL on the partitioned dataset)
disagrees, with the other three in agreement — a single dissenting
result rather than a genuine method-level split.

# Customisation

The remaining examples use the compact `sample_trees` dataset to
demonstrate optional arguments. Reload the backbone and rebuild the node
presence matrix:

```{r reload-sample}
backbone <- sample_trees[["70p_uce"]]
others   <- sample_trees[names(sample_trees) != "70p_uce"]
support_type <- c(
  "70p_ASTRAL_partition_entropy" = "lpp",
  "70p_ASTRAL_uce"               = "lpp",
  "70p_ghost"                    = "ufboot",
  "70p_partition_entropy"        = "ufboot"
)
npm <- node_presence_matrix(backbone, others, support_col = c(1, 2),
                            support_type = support_type)
```

## Universal thresholds (no support_type)
If `support_type` is not declared when building the node presence matrix, 
`plot_phylorug()` falls back to universal thresholds. Values greater than 1 are 
divided by 100 to bring them onto a 0–1 scale, and all trees are binned against 
the same cutoffs regardless of metric. This is useful for quick exploration but 
less precise than metric-specific binning:
```{r universal_threshold }
npm_universal <- node_presence_matrix(backbone, others, support_col = c(1, 2))

plot_phylorug(
  backbone, npm_universal,
  mode = "support"
)
```

## Grid dimensions

By default, phylorug chooses a roughly square grid. Override with
`n_rows` and `n_cols`:

```{r grid-demo}
#rug with flatend grid
plot_phylorug(backbone, npm, n_rows = 1, n_cols = 4)
```

```{r}
#rug with squired grid
plot_phylorug(backbone, npm, n_rows = 2, n_cols = 2)
```

## Tree appearance

`plot_phylorug()` passes additional arguments through `...` to
`ape::plot.phylo()`. Common options:

```{r tree-appearance}
plot_phylorug(backbone, npm, cex = 0.6, font = 3, edge.width = 1.5)
```

## Including the backbone

By default, the backbone does not occupy a cell in the rug (it trivially
recovers every clade, since it defines the topology). Set
`include_backbone = TRUE` to add it as cell 1:

```{r include-bb}
plot_phylorug(backbone, npm, include_backbone = TRUE)
```

In support mode, pass the backbone's support metric directly to `include_backbone` so it gets binned against the correct thresholds:
```{r}
plot_phylorug(backbone, npm,
              mode             = "support",
              include_backbone = "ufboot")
```

## Custom thresholds

Override the default support bins with your own cutoffs:

```{r custom-thresh, fig.dpi=300}
my_thresholds <- list(
  ufboot = c(very_high = 98, high = 90, moderate = 70),
  lpp    = c(very_high = 0.99, high = 0.90, moderate = 0.70)
)

plot_phylorug(
  backbone, npm,
  mode       = "support",
  thresholds = my_thresholds
)
```

# Tips and best practices

**Root and prune before translating.** `translate_tips()` replaces
specimen codes with species names. Once translated, you cannot match
outgroup names for rooting. Always: (1) root, (2) drop outgroup, (3)
translate.

**One file = one analysis.** `read_trees()` treats each file as one
analysis. A single-tree file returns a `phylo`; a file with multiple
equally optimal trees (as POY, TNT, or PAUP\* write them) returns a
`multiPhylo` scored as a pool.

**Pool scoring in presence mode.** By default (`pool_threshold = 1.0`), a clade counts as present only if every tree in the pool recovers it — equivalent to a strict consensus. Set `pool_threshold = 0.5` for majority rule, or `pool_threshold = 0` to record the raw proportion (e.g., 2 out of 3 gives 0.67). A single-tree analysis always records 1 or 0.

**Support mode is not available for pools.** Node labels on tied optima
are not independent support values. Meaningful support (bootstrap,
posterior, SH-aLRT) comes only from explicit resampling or model-based
procedures. To visualize support for a parsimony analysis, compute it
externally (e.g., map bootstrap replicates onto the strict consensus,
per Simmons and Freudenstein 2011), save the summarized tree, and pass
that single file to phylorug. Passing raw tied optima into support mode
would display arbitrary node labels as though they were support values,
producing an artifactual rug.

**Wrap bare `multiPhylo` in a list.** If you construct comparison trees
manually, each element of the list passed to `node_presence_matrix()`
must be a single `phylo` or a `multiPhylo` (a pool from one search). A
bare `multiPhylo` at the top level is interpreted as one analysis with
multiple tied-optimal trees, not as separate analyses.

**Use `prune_to_shared()` when taxa differ.** If a comparison tree is
missing some backbone taxa, `node_presence_matrix()` will refuse to
proceed (a clade containing a missing taxon cannot be evaluated). Run
`check_taxa()` first to see the discrepancies, then `prune_to_shared()`
to reduce all trees to their common taxa.

# References

- Fu, Y. et al. (2025). Phylogenomic insights into the higher level
  relationships within Culicomorpha (Diptera). *Insect Systematics and
  Diversity*, 9(6), ixaf056.
- Giribet, G. (2003). Stability in phylogenetic formulations and its
  relationship to nodal support. *Systematic Biology*, 52(4), 554–564.
- Lopes, F. et al. (2024). From museum drawer to tree: Historical DNA
  phylogenomics clarifies the systematics of rare dung beetles
  (Coleoptera: Scarabaeinae) from museum collections. *PLOS ONE*,
  19(12), e0309596.
- Montanaro, G., Lopes, F., Gunter, N.L., Scholtz, C., Davis, A.L.,
  Losacco, F., Rossini, M., Gillett, C.P.D.T., Saxton, N.A., Stone, R.L.,
  Daniel, G.M., and Tarasov, S. (2026). Phylogenomics resolves a
  200-year-old puzzle: a revised tribal classification of Afro-Eurasian
  dung beetles (Coleoptera: Scarabaeinae). *bioRxiv*.
  https://doi.org/10.64898/2026.07.22.740134
- Machado, D. J. (2015). YBYRÁ fossile: a command line system for total
  evidence dating and sensitivity analysis. *BMC Bioinformatics*, 16,
  40. 
- Sanders, K. L. (2010). Cladescan: exhaustive phylogenetic searches of
  consensus support in large data sets. *Cladistics*, 26(6), 598–613.
- Steenwyk, J. L., Li, Y., Zhou, X., Shen, X. X. & Rokas, A. (2023).
  Incongruence in the phylogenomics era. *Nature Reviews Genetics*,
  24(12), 834–850.
- Wheeler, W. C. (1995). Sequence alignment, parameter sensitivity, and
  the phylogenetic analysis of molecular data. *Systematic Biology*,
  44(3), 321–331.
